quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-09-01 paper-quollr.Rmd
Updated w_h definition
Updated elements in an object (`nldr_scaled_obj`)

This function requires several parameters: the high-dimensional data (`highd_data`), the emdedding data (`nldr_data`), the number of bins along the x-axis (`b1`), the buffer amount as a proportion of data (`q`), and benchmark value to extract high density hexagons (`benchmark_highdens`). The function returns an object that includes the scaled NLDR object (`nldr_obj`), the hexagonal object (`hb_obj`), the fitted model in both $2\text{-}D$ (`model_2d`), and $p\text{-}D$ (`model_highd`), and triangular mesh (`trimesh_data`).

@@

The configurations of a hexagonal grid are determined by the number of bins and the bin width in each direction. The function `calc_bins_y()` is used for this purpose. This function accepts (1) an object containing three elements: the first is the scaled NLDR data with two columns, and the second and third are the limits of the original NLDR data; (2) the number of bins along the x-axis (`b1`), and (3) the buffer amount as a proportion (`q`). The buffer ensures that the grid fully covers the data space by extending one hexagon width ($a_1$) and height ($a_2$) beyond the observed data in all directions. By default, $q = 0.1$, but it must be set to a value smaller than the minimum data value to avoid exceeding the data range.

@@

Points are allocated to bins based on the nearest centroid of the hexagonal bins. The hexagonal binning algorithm can be executed using the `hex_binning()` function, or its individual components can be run separately for added flexibility. While running the process step by step would involve generating centroids, constructing hexagon coordinates, assigning points to bins, standardizing counts, and mapping the data back to hexagons, the `hex_binning()` function automates this entire workflow. The parameters used within `hex_binning()` include an object containing scaled NLDR data in the first and second columns, along with numeric vectors that represent the limits of the original NLDR data (`nldr_obj`), the number of bins along the x-axis (`b1`), and the buffer amount as a proportion of the data (`q`). The output is an object of the `hex_bin_obj` class, which contains the bin widths in each direction (`a1`, `a2`), the number of bins in each direction (`bins`), the coordinates of the hexagonal grid starting point (`start_point`), the details of bin centroids (`centroids`), the coordinates of bins (`hex_poly`), NLDR components with their corresponding hexagon IDs (`data_hb_id`), hex bins with their corresponding standardized counts (`std_cts`), the total number of bins (`tot_bins`), the number of non-empty bins (`non_bins`), and the points within each hexagon (`pts_bins`).

@@

The function begins by grouping the data by hexagon ID (`h`) and counting the number of NLDR points falling within each bin. These raw counts are stored as `n_h`. To enable comparisons across bins with varying densities, the function then standardizes these counts by dividing each bin’s count by the maximum count across all bins. This yields a normalized metric, `w_h`, ranging from 0 to 1.

>>>>

This function requires several parameters: the high-dimensional data (`highd_data`), the emdedding data (`nldr_data`), the number of bins along the x-axis (`b1`), the buffer amount as a proportion of data (`q`), and benchmark value to extract high density hexagons (`benchmark_highdens`). The function returns an object (`nldr_scaled_obj`) containing three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); the hexagonal object (`hb_obj`), the fitted model in both $2\text{-}D$ (`model_2d`), and $p\text{-}D$ (`model_highd`), and triangular mesh (`trimesh_data`).

@@

The configurations of a hexagonal grid are determined by the number of bins and the bin width in each direction. The function `calc_bins_y()` is used for this purpose. This function accepts (1) an object (`nldr_scaled_obj`) containing three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); (2) the number of bins along the x-axis (`b1`), and (3) the buffer amount as a proportion (`q`). The buffer ensures that the grid fully covers the data space by extending one hexagon width ($a_1$) and height ($a_2$) beyond the observed data in all directions. By default, $q = 0.1$, but it must be set to a value smaller than the minimum data value to avoid exceeding the data range.

@@

Points are allocated to bins based on the nearest centroid of the hexagonal bins. The hexagonal binning algorithm can be executed using the `hex_binning()` function, or its individual components can be run separately for added flexibility. While running the process step by step would involve generating centroids, constructing hexagon coordinates, assigning points to bins, standardizing counts, and mapping the data back to hexagons, the `hex_binning()` function automates this entire workflow. The parameters used within `hex_binning()` include an object (`nldr_scaled_obj`) containing three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); the number of bins along the x-axis (`b1`), and the buffer amount as a proportion of the data (`q`). The output is an object of the `hex_bin_obj` class, which contains the bin widths in each direction (`a1`, `a2`), the number of bins in each direction (`bins`), the coordinates of the hexagonal grid starting point (`start_point`), the details of bin centroids (`centroids`), the coordinates of bins (`hex_poly`), NLDR components with their corresponding hexagon IDs (`data_hb_id`), hex bins with their corresponding standardized counts (`std_cts`), the total number of bins (`tot_bins`), the number of non-empty bins (`non_bins`), and the points within each hexagon (`pts_bins`).

@@

The function begins by grouping the data by hexagon ID (`h`) and counting the number of NLDR points falling within each bin. These raw counts are stored as `n_h`. To enable comparisons across bins with varying densities, the function then standardizes these counts by dividing each bin’s count by the maximum count across all bins. This yields a standardized bin counts, `w_h`, ranging from 0 to 1.
